The tracing of trade

### 2021-02-12 Manuscript/Manuscript.tex
final review

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system

@@

@@

%The analysis explores how the quantitative approach provides a useful tool for the interpretation of the economic processes.

@@

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, duncan1982economy, Chic_hispania_1997,millet_anforas_1998}. However, the complexity and sheer scale of the process makes its understanding challenging, 
%se repite despite aquí dos veces
despite several archaeological excavations and the existence of some written records discussing its structure. Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was this massive amphora production organised? Can we detect the economic ties between provinces? Was this distribution centralised or each region organised it differently?

Recent advances in Roman studies have currently led to an environment with various interpretations on these commercial dynamics~\citep{temin_economy_2006,quantifyingwilson2009}. %The application of diverse novel quantitative approaches to new accessible datasets has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the dynamics of Roman production
The application of novel quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the Roman production
\citep{brughmans_roman_2016,orengo_seeds_2016,bayesian_2018,coto-sarmiento_identifying_2018,rubio-campillo_ecology_2018}. This work aligns with these efforts and focuses on the distribution of olive oil towards Roman provinces in the borders. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in these provinces.

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence, there is still no consensus on the structure that generated this pattern; how were the economic relations between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?
%repite aims dos veces  
This paper aims to explore these questions regarding the olive oil market by identifying links between production centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces, our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas: a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be found in specific provinces because they were shipped together.
%find in or find on

@@

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centre based on the amphoric stamps found in each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to establish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

@@

The chronology of the Baetican workshops is typically defined between the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Archaeological evidence suggests that this highly specialised production was developed across a long activity period\citep{remesal_anforas_2004}. The process seemed to have been planned at some level as is observed by the relatively minor changes in amphoric typologies and the high degree of standardisation in the production.

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 type amphorae found in all the Empire and particularly in the Western provinces~\citep{dressel_ricerche_1878,millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements~\citep{berni_millet_epigrafianforica_2008}. They have a recognisable globular form with a short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure} 
%found in find in
Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found in nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the disposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. Despite the unsolved question of ownership, some interesting ideas can be explored based on the finding of similar stamps across the Empire. Observable patterns can be used to explore how this production was organised and whether it is possible to identify specific production processes for olive oil trade. Our questions here will be focused on the distribution of amphoric stamps; were different stamps shipped through different distribution routes? Were specific stamp codes used in different workshops?

@@

Some authors argue elsewhere that this same \textit{praefectura annonae} could also have organised the supply of additional goods such as olive oil to the Roman legions~\citep{remesal_annona_1986,remesal_annona_1990}. From this perspective, olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in civilian sites close to the military garrisons also suggests the existence of diverse supply dynamics involving systematic trade between military sites to civil settlements~\citep{remesal_annona_1986, carreras_britannia_1998}. An explanation for the interest for exporting olive oil to the military garrison in the Roman provinces is still unclear. The reasons behind this effort to supply olive oil to all soldiers may be related to two aspects: a) consumption based on cultural reasons (i.e. social identification or habit by a group of consumers) and b) economical and political reasons to create a redistributive system mechanism to supply military camps~\citep[69-70]{carreras_britannia_1998}.

The Roman conquest of \textit{Britannia} and \textit{Germania} was a new opportunity to expand economical activity to support military efforts. High demand for olive oil in both provinces stimulated an important trade network involving different agents and the Empire itself. \textit{Baetica} became a strategic province for the production and distribution in both provinces as can be observed by the large volume of Dressel 20 amphora vessels found both in civil and military sites of these areas during and after the conquest.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced in the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Archaeological records suggest that this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century AD onwards \citep{peacock_amphorae_1991,carreras_britannia_1998}. At this moment, we detect an increase in the olive oil exports linked to the arrival of legions caused by different military campaigns~\citep[161]{monfort_britanniaen_1998}.

@@

The consumption of Baetican olive oil would experience a progressive slowdown from the third century AD onwards. This date matches a change in market strategy in the Empire. This gradual decrease in \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by other types of amphorae from other places and particularly the African provinces~\citep{rodriguez1991aceite,millet_anforas_1998}.

@@

The distribution of Dressel 20 amphorae in the provinces of \textit{Germania Inferior} and \textit{Germania Superior} is not as well interpreted as the rest of the provinces discussed in this work~\citep[293]{remesal_baetica_2002}. 
%it is clear no se si ponerlo
It is clear that there was a massive import of Dressel 20 in both provinces as is observed in archaeological excavations across the German \textit{limes}. However, it is still unknown to what extent trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}.

In any case, the introduction of olive oil in \textit{Germania} shows similar patterns to the one discussed for\textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region, while a majority of recovered Dressel 20 amphorae are located in the military sites built over the German \textit{limes} \citep{remesal_germaniaengl_2002}. The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a major concentration at the German \textit{limes}. It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons~\citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence clarifying the meaning of the amphoric stamps.

Some studies assumed that Roman naval technology was not suited for oceanic transport and for this reason defined riverine shipping across Rh\^one and Rhine rivers as the basic supply route for the German \textit{limes} \citep{marliere2001tonneau}. However, more recent large-scale computational analysis, as well as new excavations on the Iberian Atlantic coast suggest that the Atlantic route, which was far more efficient, may have played a major role in the supply of \textit{Germania} and \textit{Britannia}~\citep{rubio-campillo_ecology_2018}.

@@

Our dataset also contained a categorical variable defining the \textit{conventus} for each production site. The \textit{conventus} were administrative centres for territorial organisation under the Roman Empire \citep[58]{ozcariz_gil_administracion_2013}. The creation of the \textit{conventus} in some Roman provinces was a form of control with the purpose of organising provincial administration~\citep{albertini_les_1923}. The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter \textit{Hispalis}), \textit{Cordubensis} (currently C\'ordoba, hereafter \textit{Corduba}) and \textit{Astigi} (currently \'Ecija, Sevilla, hereafter \textit{Astigi})~\citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

The distribution of codes in stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this number (267 and 146 stamps respectively). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exception of two large workshops in \textit{Hispalis} (La Catria and Arva). In these two workshops a comparatively large amount of stamp codes were found (29 different codes on each of them). According to previous studies, those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,arva_1997}.

@@

The analysis of stamps in consumption areas used also the CEIPAC database and filters performed for the previous dataset. However, in this case there was a large number of sites with only one or two stamps which may introduce noise in the similarity analysis. This is the reason why in this case an additional filter was performed and we only selected sites having at least five stamps.

@@

Similarity between amphora workshops and/or consumption sites was measured by quantifying a distance (also called dissimilarity) between the set of codes of stamps found for each pair of sites (i.e. to what extent the stamp codes detected in these sites were different). The chosen measure for this study was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of stamps while taking into account the frequency that each code is present on each site. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

As we previously discussed some bias on the intensity of archaeological works and workshop identification may have an impact on this analysis. Further fieldwork could provide a clearer picture of the links between workshops if they allow increasing the sample size of stamp codes is increased.
%mirar el increased que creo que está mal puesto

@@

The areas in \textit{Britannia} with most intense military activity have some degree of overlap with civil areas. However, we do not detect this overlap for the German \textit{limes} \citep{xanten2018}. This could be interpreted by parallel supply systems but it also could be caused by intensity bias as military centres have been more excavated than civil centres.
%civil centres no sé si está bien
Finally, the results suggest that there is no clear organisational pattern linking production centres and/or consumption centres beyond local spatial proximity within regions. As a consequence, our analysis does not indicate the presence of specific Baetican amphoric workshop consistently supplying specific regions or even provinces. This outcome suggests that stamp codes cannot be used to identify whether production centres had fixed clients and other proxies would be needed to identify this link in case it existed. Based on our analysis production centres could have distributed olive oil to both \textit{Britannia} and \textit{Germania} at a short-term basis of demand and without any long-term planning or centralised decision-making system.

@@

This work has used quantitative analysis to explore the structure of olive oil long-range trade using the stamps found in Dressel 20 amphora vessels. The most interesting result is linked to the meaning of these inscriptions. The use of specific codes seems to be almost exclusively decided by each workshop, so it could be used to identify amphorae found in consumption centres (i.e. an amphora with a given stamp was probably made on a specific workshop). This hypothesis is sustained by the fact that the similarity of stamps between workshops is very low and a majority of stamp codes are only found at one workshop. On the other hand, the presence of stamp codes in small groups of nearby workshops could suggest that the production of these workshops was somehow connected (e.g. they belonged to the same owner or were collectively producing a batch for the same producer).

>>>>

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{color}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system

@@

%\textcolor{red}{We have integrated all these changes and we have improved the manuscript with an extensive revision of the text by a native reader. So it means that we cannot track every single change because the manuscript has been entirely changed}

@@

@@

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades~\citep{rodriguez_economioleicola_1977, duncan1982economy, Chic_hispania_1997,millet_anforas_1998}. However, the complexity and sheer scale of the process makes its understanding challenging, despite several archaeological excavations and the existence of some written records discussing its structure. Our knowledge of this economic activity has benefited from new findings and data sources, although all these advances several questions remain open: How was this massive amphora production organised? Can we detect the economic ties between provinces? Was this distribution centralised or each region organised it differently?

Recent advances in Roman studies have currently led to an environment with various interpretations on these commercial dynamics~\citep{temin_economy_2006,quantifyingwilson2009}. 
The application of novel quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the dynamics of Roman production~\citep{brughmans_roman_2016,orengo_seeds_2016,bayesian_2018,coto-sarmiento_identifying_2018,rubio-campillo_ecology_2018}. This work aligns with these efforts and focuses on the distribution of olive oil towards Roman provinces in the borders. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in these provinces.

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence, there is still no consensus on the structure that generated this pattern; how were the economic relations between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?

This paper proposes to explore these questions regarding the olive oil market by identifying links between production centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces, our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas: a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be detected in specific provinces because they were shipped together.

@@

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centres based on the amphoric stamps found in each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to establish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

@@

The chronology of the Baetican workshops is typically defined between the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Archaeological evidence suggests that this highly specialised production was developed across a long activity period~\citep{remesal_anforas_2004}. The process seemed to have been planned at some level as is observed by the relatively minor changes in amphoric typologies and the high degree of standardisation in the production.

The importance of this olive oil trade is revealed by the massive number of Dressel 20 type amphorae found in all the Empire and particularly in the Western provinces~\citep{dressel_ricerche_1878,millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civilian settlements~\citep{berni_millet_epigrafianforica_2008}. They have a recognisable globular form with a short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout the Empire~\citep{berni_dressel_2016}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure}

Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found in nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the disposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves~ \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. Despite the unsolved question of ownership, some interesting ideas can be explored based on the finding of similar stamps across the Empire. Observable patterns can be used to explore how this production was organised and whether it is possible to identify specific production processes for olive oil trade. Our questions here will be focused on the distribution of amphoric stamps; were different stamps shipped through different distribution routes? Were specific stamp codes used in different workshops?

@@

Some authors argue elsewhere that this same \textit{praefectura annonae} could also have organised the supply of additional goods such as olive oil to the Roman legions~\citep{remesal_annona_1986,remesal_annona_1990}. From this perspective, olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in civilian sites close to the military garrisons also suggests the existence of diverse supply dynamics involving systematic trade between military sites to civilian settlements~\citep{remesal_annona_1986, carreras_britannia_1998}. An explanation for the interest for exporting olive oil to the military garrison in the Roman provinces is still unclear. The reasons behind this effort to supply olive oil to all soldiers may be related to two aspects: a) consumption based on cultural reasons (i.e. social identification or habit by a group of consumers) and b) economical and political reasons to create a redistributive system mechanism to supply military camps~\citep[69-70]{carreras_britannia_1998}.

The Roman conquest of \textit{Britannia} and \textit{Germania} was a new opportunity to expand economical activity to support military efforts. High demand for olive oil in both provinces stimulated an important trade network involving different agents and the Empire itself. \textit{Baetica} became a strategic province for the production and distribution in both provinces as can be observed by the large volume of Dressel 20 amphora vessels found both in civilian and military sites of these areas during and after the conquest.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced in the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Archaeological records suggest that this absence of olive oil changed with the arrival of a large number of Dressel 20 amphorae to \textit{Britannia} from the first century AD onwards \citep{peacock_amphorae_1991,carreras_britannia_1998}. At this moment, we detect an increase in the olive oil exports linked to the arrival of legions caused by different military campaigns~\citep[161]{monfort_britanniaen_1998}.

@@

The consumption of Baetican olive oil would experience a progressive slowdown from the third century AD onwards. This date matches a change in market strategy in the Empire. This gradual decrease in \textit{Baetican} olive oil exports can be observed as a decrease in the number of Dressel 20 detected in excavated sites as this typology is replaced by other types of amphorae from other places and particularly the African provinces~\citep{rodriguez1991aceite,millet_anforas_1998}.

@@

The distribution of Dressel 20 amphorae in the provinces of \textit{Germania Inferior} and \textit{Germania Superior} is not as well interpreted as the rest of the provinces discussed in this work~\citep[293]{remesal_baetica_2002}.

It is clear that there was a massive import of Dressel 20 in both provinces as is observed in archaeological excavations across the German \textit{limes}. However, it is still unknown to what extent trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons~\citep[156]{remesal_germn_2010}.

In any case, the introduction of olive oil in \textit{Germania} shows similar patterns to the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region, while a majority of recovered Dressel 20 amphorae are located in the military sites built over the German \textit{limes}~\citep{remesal_germaniaengl_2002}. The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civilian settlement and military sites with a major concentration at the German \textit{limes}. It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons~\citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence clarifying the meaning of the amphoric stamps.

Some studies assumed that Roman naval technology was not suited for oceanic transport and for this reason defined riverine shipping across Rh\^one and Rhine rivers as the basic supply route for the German \textit{limes}~\citep{marliere2001tonneau}. However, more recent large-scale computational analysis, as well as new excavations on the Iberian Atlantic coast suggest that the Atlantic route, which was far more efficient, may have played a major role in the supply of \textit{Germania} and \textit{Britannia}~\citep{rubio-campillo_ecology_2018}.

@@

Our dataset also contained a categorical variable defining the \textit{conventus} for each production site. The \textit{conventus} were administrative centres for territorial organisation under the Roman Empire~\citep[58]{ozcariz_gil_administracion_2013}. The creation of the \textit{conventus} in some Roman provinces was a form of control with the purpose of organising provincial administration~\citep{albertini_les_1923}. The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter \textit{Hispalis}), \textit{Cordubensis} (currently C\'ordoba, hereafter \textit{Corduba}) and \textit{Astigi} (currently \'Ecija, Sevilla, hereafter \textit{Astigi})~\citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

The distribution of codes in stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this number (267 and 146 stamps respectively). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exception of two large workshops in \textit{Hispalis} (La Catria and Arva). In these two workshops a comparatively large number of stamp codes were found (29 different codes on each of them). According to previous studies, those workshops became the most important centres of amphora production in the region~\citep{rodriguez_economioleicola_1977,arva_1997}.

@@

The analysis of stamps in consumption areas also used the CEIPAC database and filters performed for the previous dataset. However, in this case there was a large number of sites with only one or two stamps which may introduce noise in the similarity analysis. This is the reason why in this case an additional filter was performed and we only selected sites having at least five stamps.

@@

Similarity between amphora workshops and/or consumption sites was measured by quantifying a distance (also called dissimilarity) between the set of codes of stamps found for each pair of sites (i.e. to what extent the stamp codes detected in these sites were different). The chosen measure for this study was the Morisita-Horn index~\citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of stamps while taking into account the frequency that each code is present on each site. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

As we previously discussed some bias on the intensity of archaeological works and workshop identification may have an impact on this analysis. Further fieldwork could provide a clearer picture of the links between workshops if these contributions increase the sample size of the stamp code dataset.

@@

The areas in \textit{Britannia} with most intense military activity have some degree of overlap with civilian areas. However, we do not detect this overlap for the German \textit{limes} \citep{xanten2018}. This could be interpreted by parallel supply systems but it also could be caused by intensity bias as military centres have been more excavated than civilian centres.

Finally, the results suggest that there is no clear organisational pattern linking production centres and/or consumption centres beyond local spatial proximity within regions. As a consequence, our analysis does not indicate the presence of specific Baetican amphoric workshop consistently supplying specific regions or even provinces. This outcome suggests that stamp codes cannot be used to identify whether production centres had fixed clients and other proxies would be needed to identify this link in case it existed. Based on our analysis production centres could have distributed olive oil to both \textit{Britannia} and \textit{Germania} at a short-term basis of demand and without any long-term planning or centralised decision-making system.

@@

This work has used quantitative analysis to explore the structure of olive oil long-range trade using the stamps found in Dressel 20 amphora vessels. The most interesting result is linked to the meaning of these inscriptions. The use of specific codes seems to be almost exclusively decided by each workshop, so it could be used to identify amphorae found in consumption centres (i.e. an amphora with a given stamp was probably made on a specific workshop). This hypothesis is sustained by the fact that the similarity of stamps between workshops is very low and a majority of stamp codes are only detected in one workshop. On the other hand, the presence of stamp codes in small groups of nearby workshops could suggest that the production of these workshops was somehow connected (e.g. they belonged to the same owner or were collectively producing a batch for the same producer).
